---
title: "Carbon and environmental valuation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Carbon and environmental valuation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(greenbook)
```

## Carbon values for appraisal

DESNZ (formerly BEIS) publishes a series of marginal abatement
cost values out to 2100 with low / central / high scenarios for
both the traded and non-traded sectors. `gb_carbon_value()` looks
them up:

```{r}
gb_carbon_value(2024)
gb_carbon_value(2024, scenario = "high")
gb_carbon_value(2030)
```

The values are published in 2020 prices. Rebase to a different
year via `base_year`:

```{r}
gb_carbon_value(2030, base_year = 2024)
```

## Net present value of an emissions path

A greenhouse-gas-emitting project has its environmental
externality monetised at the published carbon value, then
discounted under the standard STPR.

```{r}
# 100 tCO2e per year for 2024 to 2030, treated as a cost
emissions <- rep(100, 7)
years <- 2024:2030
gb_carbon_npv(emissions, years, base_year = 2024)
```

For an abatement project (avoided emissions), use
`sign = "benefit"`:

```{r}
gb_carbon_npv(emissions, years, base_year = 2024, sign = "benefit")
```

## Wellbeing valuation

Life-satisfaction changes can be monetised via WELLBYs (HMT
Wellbeing Guidance for Appraisal, July 2021). Default unit value:
GBP 13,000 per WELLBY in 2019 prices.

```{r}
# 1-point lift in life satisfaction, 100 people, 5 years
gb_wellby(1, persons = 100, years = 5)

# In 2024 prices
gb_wellby(1, persons = 100, years = 5, base_year = 2024)

# Sensitivity range
gb_wellby(1, persons = 100, years = 5, scenario = "low")
gb_wellby(1, persons = 100, years = 5, scenario = "high")
```

## Risk to life and health

Value of Preventing a Fatality (VPF) and Quality-Adjusted Life
Year (QALY) per DfT TAG and DHSC supplementary guidance.

```{r}
gb_vpf()        # 2024 published value: GBP 2.153m
gb_vpf(2018)    # 2018 baseline

gb_qaly(1)                          # DHSC: GBP 70k per QALY
gb_qaly(1, scenario = "nice_upper") # NICE upper threshold: GBP 30k
```

## Provenance

```{r}
gb_data_versions()
```

## Limitations

The bundled DESNZ carbon path is a stub vintage for v0.1.0 (annual
values 2020 to 2050, scenarios linearly interpolated). Future
releases will refresh from the full DESNZ workbook including
2050 to 2100 horizons.

The VPF series uses a 2 percent annual real uplift between
DfT-published anchors (2018 and 2024) as a proxy for real GDP per
head growth. For high-precision applications, consult the latest
TAG data book directly.
